---
title: "NYPD_Shooting_Report"
author: "Wojciech Gajewski"
date: "2024-08-15"
output: pdf_document
---

## Load the dataset into R:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(caret)
library(lubridate)
library(randomForest)

file_path <- "C:\\Users\\wgaje\\Documents\\R_assaiment\\NYPD_Shooting_Incident_Data__Historic__2024.csv"
df <- read.csv(file_path)

str(df)
summary(df)
```

## Tidy the Data

Perform data cleaning and tidying to make the dataset more manageable:

```{r dataprocessing}
df <- df %>%
  select(INCIDENT_KEY, OCCUR_DATE, OCCUR_TIME, BORO, LOC_OF_OCCUR_DESC, PRECINCT, Latitude, Longitude, STATISTICAL_MURDER_FLAG)

# Convert OCCUR_DATE to Date format and OCCUR_TIME to Time format
df <- df %>%
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, format="%m/%d/%Y"),
         OCCUR_TIME = hms::as_hms(OCCUR_TIME))

# Check for missing values
sum(is.na(df))

# Remove rows with missing values
df <- df %>% drop_na()
head(df)
```

### Visualization 1: Number of Incidents by Borough

```{r plotting1, echo=FALSE}
# Plot the number of incidents by borough
ggplot(df, aes(x = BORO)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Shooting Incidents by Borough", x = "Borough", y = "Number of Incidents") +
  theme_minimal()

```

### Visualization 2: Shooting Incidents Over Time

```{r plotting2, echo=FALSE}
# Plot incidents over time
ggplot(df, aes(x = OCCUR_DATE)) +
  geom_histogram(binwidth = 30, fill = "darkred", color = "white") +
  labs(title = "Shooting Incidents Over Time", x = "Date", y = "Number of Incidents") +
  theme_minimal()
```

## Build a Simple Model

We'll build a simple model to predict whether an incident was a murder or not based on the available features.

### Model: Random Forest

```{r modeling}
# Convert STATISTICAL_MURDER_FLAG to a factor
df$STATISTICAL_MURDER_FLAG <- as.factor(df$STATISTICAL_MURDER_FLAG)

set.seed(123)
train_index <- createDataPartition(df$STATISTICAL_MURDER_FLAG, p = 0.7, list = FALSE)
train_data <- df[train_index, ]
test_data <- df[-train_index, ]

colSums(is.na(train_data))
preProcess_missingdata_model <- preProcess(train_data, method = 'medianImpute')
train_data <- train_data[, colSums(is.na(train_data)) == 0]

rf_model <- randomForest(STATISTICAL_MURDER_FLAG ~ ., data = train_data, importance = TRUE)
predictions <- predict(rf_model, newdata = test_data)

confusion_matrix <- confusionMatrix(predictions, test_data$STATISTICAL_MURDER_FLAG)
print(confusion_matrix)
```

## Summary of Bias in the Model:

### Accuracy and NIR:

The model's accuracy (80.02%) is close to the No Information Rate (80.63%), suggesting it mostly predicts the majority class ("false" for non-murder).

### Kappa Statistic:
A low Kappa (0.1241) indicates poor performance beyond random chance, highlighting the model's difficulty in handling class imbalance.

### Sensitivity and Specificity:

Sensitivity (96.14% for "false"): The model is highly sensitive to predicting non-murders. Specificity (12.92% for "true"): The model struggles to correctly identify murders, reflecting bias towards the majority class.

### Bias Implications:

The model is biased towards predicting the majority class (non-murders), leading to poor detection of the minority class (murders). This imbalance is critical, especially in safety-related predictions, where identifying murders accurately is crucial.
